# More Random Effects

Previously we've looked at random intercepts, but any observation level covariate effect could be allowed to vary by cluster as well.


## Application

Returning to the GPA data, recall the visualization from before.


```{r spaghetti2, echo=FALSE}
set.seed(1234)
gpa_lm = lm(gpa ~ occasion, data=gpa)
sample_students = gpa %>% filter(student %in% sample(1:200, 10))
occasion_sample = gpa$occasion[gpa$student %in% sample_students$student]
gpa_sample = gpa$gpa[gpa$student %in% sample_students$student]
gpa %>% 
  modelr::add_predictions(gpa_lm, var='all') %>% 
  group_by(student) %>% 
  plot_ly %>% 
  add_lines(x=~occasion, y=~gpa, opacity=.2, color=I(palettes$orange$complementary), showlegend=F) %>%
  add_paths(x=occasion_sample, y=gpa_sample, color=I(palettes$orange$orange), opacity=.5, showlegend=F) %>% 
  add_lines(x=~occasion, y=~all, color=I('#b2001d')) %>% 
  theme_plotly()
```

<br>
Let us now assume that the trend over time is allowed to vary by student.  Using <span class="pack">lme4</span>, this is quite straightforward.

```{r random_slope, eval=FALSE}
gpa_mixed =  lmer(gpa ~ occasion + (1 + occasion|student), data=gpa)
summary(gpa_mixed)
```

Pretty easy huh? Let's look at the results.


```{r random_slope_summary, echo=FALSE}
gpa_mixed =  lmer(gpa ~ occasion + (1 + occasion|student), data=gpa)
pander(tidy(gpa_mixed, 'fixed', conf.int=T) %>% mutate_if(is.numeric, arm::fround, digits=2))
tidy(VarCorr(gpa_mixed)) %>% 
  slice(-3) %>%  
  select(-var2) %>% 
  rename(variance=vcov, sd=sdcor) %>%  
  mutate_if(is.numeric, arm::fround, digits=3) %>% 
  mutate_all(function(x) ifelse(is.na(x), '', x)) %>%
  pander()
```

Note that since we have 0 as our starting semester, the intercept tells us what the average GPA is in the first semester.  The associated intercept variance tells us how much that starting GPA bounces around from student to student.

The slope variance for occasion might not look like much in comparison, but slopes are on a notably different scale than the intercept.  Note that the mean slope for the semester to semester effect, our fixed effect, is `r round(fixef(gpa_mixed)[2], 2)`, but from student to student it bounces around half that.  Thus we could expect most students to fall somewhere between a flat effect of zero to more than double the population average[^sdslopes]. 

Yet another point of interest is the correlation of the intercepts and slopes. In this case it's `r tidy(VarCorr(gpa_mixed)) %>% slice(3) %>% select(sdcor) %>%  round(2)`. That's pretty small, but the interpretation is the same as with any correlation.  In this case specifically, it tells us that those with lower intercepts would be associated with increased time trajectories.   This makes intuitive sense in that people are improving in general, and those at the bottom would have more room to improve.  However, this is very slight, and practically speaking we might not put too much weight on it.


### Comparison to many regressions


Let's compare these results to the ones we would have gotten had we run a separate regression for each student.  In what follows we see the distribution of of the estimated intercept and slope coefficients for all the students.

```{r ranints_vs_separateints, echo=FALSE}
gint = data_frame(Mixed=coef(gpa_mixed)$student[,1], Separate=gpa_lm_by_group[,1]) %>% 
  gather(key=Model, value=Intercept) %>% 
  ggplot(aes(x=Intercept)) +
  geom_density(aes(color=Model, fill=Model), alpha=.25) +
  scale_fill_manual(values=c(palettes$orange$orange, palettes$orange$complementary)) +
  ggtitle('Intercepts') +
  labs(x='', y='') +
  xlim(c(1.5,4)) +
  theme_trueMinimal() +
  theme(
    legend.key.size=unit(2, 'mm'),
    legend.title=element_text(size=8),
    legend.text=element_text(size=8),
    legend.box.spacing=unit(0, 'in'),
    legend.position=c(.75,.75))
gslopes = data_frame(Mixed=coef(gpa_mixed)$student[,2], Separate=gpa_lm_by_group[,2]) %>% 
  gather(key=Model, value=Occasion) %>% 
  ggplot(aes(x=Occasion)) +
  geom_density(aes(color=Model, fill=Model), alpha=.25, show.legend=F) +
  scale_fill_manual(values=c(palettes$orange$orange, palettes$orange$complementary)) +
  ggtitle('Slopes for occasion') +
  labs(x='', y='') +
  xlim(c(-.2,.4)) +
  theme_trueMinimal() 


gridExtra::grid.arrange(gint, gslopes, ncol=2)
```

Here we can see that the mixed model intercepts are generally not as extreme, i.e. the tails of the distribution have been pulled toward the overall effect.  Same goes for the slopes. In both cases the mixed model shrinks what otherwise would have been the by-group estimate, which would overfit.  This <span class="emph">regularizing</span> effect is yet another bonus when using mixed models.


### Visual prediction

Let's compare our results visually. First there is the linear regression fit. We assume the same trend for everyone.  If we add the conditional predictions that include the subject specific effects from the mixed model, we now can also make subject specific predictions, greatly enhancing the practical use of the model.  As the code to create this plot (using <span class="pack">plotly</span>) was very easy, I go ahead and show it.

```{r visualize_mixed_fit, echo=-1, eval=-1}
going_down_now = factor(rep(coef(gpa_mixed)$student[,'occasion']<0, e=6), labels=c('Up', 'Down'))
gpa %>% 
  modelr::add_predictions(gpa_lm, var='lm') %>% 
  modelr::add_predictions(gpa_mixed, var='mixed') %>% 
  group_by(student) %>% 
  plot_ly %>% 
  add_lines(x=~occasion, y=~lm, opacity=1, color=I('#ff5500'), name='Standard\nRegression') %>%
  add_lines(x=~occasion, y=~mixed, opacity=.2, color=I('#03b3ff'), name='Mixed\nModel') %>%
  theme_plotly()
```

By contrast, the by-group approach is more noisy due to treating everyone independently.  Many more students are expected to have downward or flat trends relative to the mixed model (the mixed model only had `r sum(coef(gpa_mixed)$student[,'occasion']<0)` trends estimated to be negative).

```{r visualize_bygroup_fit, echo=FALSE}
gpa_lm_fits_by_group = gpa %>% 
  split(.$student) %>% 
  map(~lm(gpa ~ occasion, data=.x)) %>% 
  map(fitted) %>% 
  unlist
going_down_now = factor(rep(gpa_lm_by_group[,'occasion']<0, e=6), labels=c('Up', 'Down'))
gpa %>% 
  modelr::add_predictions(gpa_lm, var='lm') %>% 
  mutate(stufit=gpa_lm_fits_by_group) %>% 
  group_by(student) %>% 
  plot_ly %>% 
  add_lines(x=~occasion, y=~lm, opacity=1, color=I(palettes$orange$orange), name='Standard\nRegression') %>%
  add_lines(x=~occasion, y=~stufit, color=~going_down_now, opacity=.2, color=I(palettes$orange$complementary), name='By Group') %>%
  theme_plotly()
```


## Exercises

#### Sleep revisited

Run the sleep study model with random coefficient for the Days effect, and interpret the results.  What is the correlation between the intercept and Days random effects?  Use the <span class="func">ranef</span> and <span class="func">coef</span> functions on the model you've created. What do you see?

```{r sleepstudy2}
library(lme4)
data("sleepstudy")
```

In the following replace <span class="objclass">model</span> with the name of your model object. Run each line, inspecting the result of each as you go along. 

```{r, eval=FALSE}
re = ranef(model)$Subject
fe = fixef(model)
apply(re, 1, function(x) x + fe) %>% t
```

The above code adds the fixed effects to each row of the random effects (the <span class="func">t</span> just transposes the result). What is the result compared to what you saw before?



#### Simulation revisited

The following shows a simplified way to simulate some random slopes, but otherwise is the same as the simulation before.  Go ahead and run the code.

```{r simSlopes, eval=FALSE}
set.seed(1234)  # this will allow you to exactly duplicate your result
Ngroups = 50
NperGroup = 3
N = Ngroups*NperGroup
groups = factor(rep(1:Ngroups, each=NperGroup))
re_int = rnorm(Ngroups, sd=.75)
re_slope = rnorm(Ngroups, sd=.25)
e = rnorm(N, sd=.25)
x = rnorm(N)
y = (2 + re_int[groups]) + (.5 + re_slope[groups])*x +  + e

d = data.frame(x, y, groups)
```

This next bit of code shows a way to run a mixed model while specifying that there is no correlation between intercepts and slopes.  There is generally no reason to do this unless the study design warrants it[^nocorr], but you could do it as a step in the model-building process, such that you fit a model with no correlation, then one with it.

```{r simSlopes2, eval=FALSE}
model_base = lmer(y ~ x + (1|groups), data=d)
model = lmer(y ~ x + (1|groups) + (0 + x|groups), data=d)
summary(model)
confint(model)



library(ggplot2)
ggplot(aes(x, y), data=d) +
  geom_point()
```

Compare model fit using the <span class="func">AIC</span> function, e.g. `AIC(model)`.  The model with the lower AIC is the better model, so which would you choose?



[^sdslopes]: In case it's not clear, I'm using the fact that we assume a normal distribution for the random effect of occasion.  A quick rule of thumb for a normal distribution is that 95% fall between $\pm$ 2 standard deviations of the mean.

[^nocorr]: I personally have not come across a situation where I'd do this in practice.  Even if the simpler model with no correlation was a slightly better fit, there isn't much to be gained by it.