---
title: "Getting Started"
output:
  distill::distill_article:
    self_contained: true
    toc: true
    css: [../css/book.css, ../css/standard_html.css]
# bibliography: cites.bib
# nocite:  | 
#   @whatever
---


```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(
  # code
  echo      = T,
  eval      = F,
  message   = F,
  warning   = F,
  error     = F,
  comment   = NA
)
```

### Data

Load the GPA data to get things started.

```{r}
load('data/gpa.RData')
```

If you want to load all data so you don't have to worry with it again (we will only use 3 or 4), uncomment and run the following:

```{r}
sapply(paste0('data/', dir('data', pattern = 'RData')), load, .GlobalEnv)
```


### Packages

The main package is `lme4`. If you need to install `lme4` first, run the following:

```{r}
install.packages('lme4')
```

Note that you also have `nlme` and `mgcv` for mixed models, both of which come with the base R installation.

Other packages you can use that will potentially be demonstrated.

```{r}
install.packages('glmmTMB')
install.packages('brms')  # may take a bit of time
```

Other packages you might find useful.

```{r}
install.packages('merTools')     # used in the document for plotting random effects
install.packages('performance')  # icc, r-squared, other stuff
```


You can install my `mixedup` package from GitHub with `remotes`. It has some nice summary and extraction features that are consistent across several packages that can do mixed models.  Use the second approach if you don't already have `rstanarm` or `brms` (they aren't required to use in general).

```{r}
# install.pacakges(remotes) # if you don't have it

remotes::install_github('m-clark/mixedup')

# if you don't already have rstanarm and/or brms

withr::with_envvar(c(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true"), 
  remotes::install_github('m-clark/mixedup')
)
```


### Example code

You will have plenty of opportunity for coding mixed models, but here is an example of the functions to be familiar with.  They will also work for `nlme`, `glmmTMB`, `brms`, and possibly others.

```{r}
library(lme4)

data(Orthodont, package = "nlme")

test_mod = lmer(distance ~ age + (age | Subject), data = Orthodont)

summary(test_mod, cor = FALSE) # don't need correlation of fixed effects

fixef(test_mod)

ranef(test_mod)

VarCorr(test_mod)
```

